|

### 2020-12-08 article/article.Rmd
final revision

---
title:  |
        | The Effect of Legislature Size on Public Spending
        | A Meta-Analysis
author: 
  - Huzeyfe Alptekin[^huzeyfe]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
  - Catarina Roman[^catarina]
date: \today
thanks: "The authors thank Guilherme Duarte, Robert Myles McDonnell, and David Skarbek for their constructive feedback. We also thank Cedric Antunes, Luis Castro, Giovanna França, Julia Oriente, and Lucas Mingardi for their excellent research assistance. Replication materials are available at <https://github.com/danilofreire/distributive-politics-meta-analysis>. We kindly acknowledge funding from the São Paulo State Science Foundation (FAPESP grant number 2018/00646-1)."
abstract: "In a seminal article, @weingast1981political argue that there is a positive relationship between legislature size and inefficiency in public expenditures. Their proposition is currently known as the \"law of $1/n$\" and has been widely debated in political science and public administration. However, recent studies have questioned the validity of the theory. In this letter, we estimate the first meta-analysis of the relationship between the number of legislators and public spending. Based on a sample of 29 articles, we find no robust evidence for the effect of legislature size on government budgets. Yet the aggregate results show significant heterogeneity. While earlier studies provide moderate support for the \"law of $1/n$\", papers using modern causal inference methods consistently find a negative relationship between seats and spending. The available evidence also indicates that proportional representation and mixed voting systems are no more likely to overspend than majoritarian ones."
abstractspacing: double
keywords: distributive politics; law of $1/n$; legislature size; meta-analysis; public spending
jelcodes: H21; H23; H50; H61
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Aggregate results show that legislative size has no significant impact on public
spending. However, the meta-regressions suggest that our study sample has high
levels of heterogeneity, and effects differ substantially according to study
specifications. A reliable predictor of effect size is the research method
employed in the original paper. While papers that employ conventional linear
regressions often find a positive relation between legislature size and
government budgets, recent studies using new methods of causal inference,
specially those with regression discontinuity designs, provide consistent
evidence for a negative relation. This goes in contrast with the original theory
and supports the existence of a "reverse law of $1/n$". Additionally, we find
that non-majoritarian voting systems decrease government spending as a
percentage of GDP, which is also evidence against the theoretical expectations
of the "law of $1/n$". In summary, we do not find support for the theory as
originally conceived, and causal inference models suggest that larger
legislatures are associated with lower public spending.

@@

Our study sample reflects the development of the literature. Although the "law
of $1/n$" was first formulated in 1981, the empirical assessment of the theory
only started a few years later, as dates of publishing range from `r min(dat$year)`
to `r max(dat$year)`. Most studies focus on the United States (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='USA')])`), but our sample also contains
papers on Australia (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='AUS')])`), Brazil (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='BRA')])`), Germany (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='DEU')])`), Indonesia (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='IDN')])`), Italy (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='ITA')])`), and
Switzerland (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='CHE')])`). Five articles use cross-national data and analyse from 2 to 110
countries. Early studies employed OLS and panel data methods to analyse the
results, whereas the use of regression discontinuity designs is a recent
addition to the literature, starting in 2012.

@@

The "law of $1/n$" states that more legislators increase government expenditure.
Here we employ three methods to assess the empirical validity of that
relationship. First, we run a binomial one-proportion z-test to determine
whether the study coefficients have a positive or negative sign. Then, we fit 7
random-effect models using the \texttt{meta} \citep{balduzzi2019perform} and the
\texttt{dmetar} \citep{dmetar2019} packages for the \texttt{R} statistical
language \citep{rstats2019}. We estimate the true effect size variance
($\tau^2$) with a Restricted Maximum Likelihood Estimator, which the literature
considers the most precise for continuous dependent variables
\citep{veroniki2016methods}. Lastly, we run a series of meta-regressions to test
possible sources of result heterogeneity. To recapitulate, our independent
variables of interest are lower chamber size ($n$), the natural logarithm of
lower chamber size (log($n$)), and upper chamber size ($k$). The dependent
variables are public expenditure per capita, the natural logarithm of public
expenditure per capita, and government expenditure as a percentage of GDP. Since
the outcomes have different scales, we treat them separately in our models.

@@

We then proceed to the meta-analysis. We matched the house size variables with
our measures of government spending and created a theoretical $3 \times 3$
matrix. Out of the 9 possible variable combinations, we found only 7 in
the article pool. Our sample includes no papers that analyse the relationships
between log($n$), and public expenditure per capita and between $k$ and the
logarithm of public expenditure per capita.

Figure \ref{fig:plots} shows the forest plots for our restricted sample, the one
which includes only the main `r dim(dat)[1]` coefficients for the 
`r length(unique(dat$id))` selected papers. On the left side of the plots are 
the name of the study authors and the paper publication year. For unpublished
studies, we included the first year the paper was available online. The bars in
the middle show the reported effect sizes and the vertical lines indicate their
average, weighted by standard errors. The length of the lines represent the
precision of the estimates. The red line at the bottom of the figures displays
the aggregate coefficients plus their respective confidence intervals.

@@

@@

\begin{landscape}
\begin{figure}[p]
\begin{center}
\caption{Forest plots of the relationship between legislature size and government spending (reduced sample)}
\vspace{.3cm}
\includegraphics[width=25cm,height=17cm]{../graphs/graph1.pdf}
\label{fig:plots}
\end{center}
\end{figure}
\end{landscape}

@@

In a nutshell, we find no robust evidence in favour of the "law of $1/n$".
While some models do show a positive and statistically significant result, none
of the prediction intervals are totally positive or negative. The studies also
have considerable heterogeneity, what indicates that the original coefficients
do not point consistently towards the same direction.

One reason why we have not detected any strong effects may be due to the
identification strategy authors use in their models. On the one hand, OLS and
panel data models require too many controls to make units comparable, and they
are also vulnerable to omitted variable bias or post-treatment bias
[@cinelli2020making; @pearl2015conditioning]. On the other hand, estimation
methods such as instrumental variables and, in particular, regression
discontinuity designs have become popular because of their internal validity
[@angrist2008mostly]. Figure \ref{fig:plots2} shows the disaggregated effects for
two sets of models that employ causal estimation techniques. They measure the
impact of lower house size on expenditure per capita (left) and on the natural
logarithm of expenditure per capita (right).

@@

\begin{figure}[htb!]
\begin{center}
\caption{Forest plots of the relationship between legislature size and government spending with regression method heterogeneity (restricted sample)}
\vspace{.3cm}
\includegraphics[width=1\linewidth]{../graphs/graph2.pdf}
\label{fig:plots2}
\end{center}
\end{figure}

Papers that employed instrumental variables show an interesting variation: out
of the four coefficients listed, two are positive and two are negative. Two of
these four coefficients are significant, again one positive and one negative.
The subgroup analysis presents a SMD which is negative but insignificant. In
contrast, all papers that use regression discontinuity designs show negative and
statistically significant results. Since only three papers in our sample use
RDDs, we are cautious about predicting an overall negative relation, but they do
indicate that better identification strategies yield a zero-to-negative impact
of legislature size on expenditure, in support of the reverse "law of $1/n$".

@@

The meta-regressions indicate that study characteristics have a considerable
influence on reported results. Factors such as the electoral system affects the
relationship between legislature size and public expenditure, but the results
are not replicable in all estimations. Publication year also produces
conflicting findings in our models. Nevertheless, the meta-regressions confirm
that RDDs produce negative effects more frequently than OLS regressions.

>>>>

---
title:  |
        | The Effect of Legislature Size on Public Spending
        | A Meta-Analysis
author: 
  - Huzeyfe Alptekin[^huzeyfe]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
  - Catarina Roman[^catarina]
date: \today
thanks: "The authors thank Guilherme Duarte, Robert Myles McDonnell, and David Skarbek for their constructive feedback. We also thank Cedric Antunes, Luis Castro, Giovanna França, Julia Oriente, and Lucas Mingardi for their excellent research assistance. Replication materials are available at <https://github.com/danilofreire/distributive-politics-meta-analysis>. We kindly acknowledge funding from the São Paulo State Science Foundation (FAPESP grant number 2018/00646-1)."
abstract: "In a seminal article, @weingast1981political argue that there is a positive relationship between legislature size and inefficiency in public expenditures. Their proposition is currently known as the \"law of $1/n$\" and has been widely debated in political science and public administration. However, recent studies have questioned the validity of the theory. In this letter, we estimate the first meta-analysis of the relationship between the number of legislators and public spending. Based on a sample of 29 articles, we find no robust evidence for the effect of legislature size on government budgets. Yet the aggregate results show significant heterogeneity. While earlier studies provide moderate support for the \"law of $1/n$\", papers using causal inference methods consistently find a negative relationship between seats and spending. The available evidence also indicates that proportional representation and mixed voting systems are no more likely to overspend than majoritarian ones."
abstractspacing: double
keywords: distributive politics; law of $1/n$; legislature size; meta-analysis; public spending
jelcodes: H21; H23; H50; H61
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Aggregate results show that legislative size has no significant impact on
public spending. However, the meta-regressions suggest that our study sample
has high levels of heterogeneity, and effects differ substantially according to
study specifications. We find that one reliable predictor of effect size is the
research method employed in the original paper. While papers that employ
conventional linear regressions often find a positive relation between
legislature size and government budgets, recent studies using modern methods of
causal inference, specially those with regression discontinuity designs,
provide consistent evidence for a negative relation. This goes in contrast with
the original theory and supports the existence of a "reverse law of $1/n$".
Additionally, we find that non-majoritarian voting systems decrease government
spending as a percentage of GDP, which is also evidence against the theoretical
expectations of the "law of $1/n$", and they do not have a relevant impact on
the other measures of public expenditure. In summary, our analysis does not
provide robust support for the theory as originally conceived, and causal
inference models suggest that larger legislatures are associated with lower
public spending.

@@

Our study sample reflects the development of the literature. Although the "law
of $1/n$" was first formulated in 1981, the empirical assessment of the theory
only started a few years later, as dates of publishing range from `r min(dat$year)`
to `r max(dat$year)`. Most studies focus on the United States (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='USA')])`), but our sample also contains
papers on Australia (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='AUS')])`), Brazil (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='BRA')])`), Germany (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='DEU')])`), Indonesia (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='IDN')])`), Italy (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='ITA')])`), and
Switzerland (`r as.numeric(table(unique(dat[,c('id','locationISO')])$locationISO)[which(names(table(unique(dat[,c('id','locationISO')])$locationISO))=='CHE')])`). Five articles use cross-national data and analyse from 2 to 110
countries. Early studies used OLS and panel data methods to estimate the
results, whereas the use of regression discontinuity designs is a recent
addition to the literature, as the first paper to apply it was published in 2012.

@@

The "law of $1/n$" states that more legislators increase government expenditure.
Here we employ three methods to assess the empirical validity of that
relationship. First, we run a binomial one-proportion z-test to determine
whether the study coefficients have a positive or negative sign. Then, we fit 9
random-effect models using the \texttt{meta} \citep{balduzzi2019perform} and the
\texttt{dmetar} \citep{dmetar2019} packages for the \texttt{R} statistical
language \citep{rstats2019}. We estimate the true effect size variance
($\tau^2$) with a Restricted Maximum Likelihood Estimator, which the literature
considers the most precise for continuous dependent variables
\citep{veroniki2016methods}. Lastly, we run a series of meta-regressions to test
possible sources of result heterogeneity. To recapitulate, our independent
variables of interest are lower chamber size ($n$), the natural logarithm of
lower chamber size (log($n$)), and upper chamber size ($k$). The dependent
variables are public expenditure per capita, the natural logarithm of public
expenditure per capita, and government expenditure as a percentage of GDP. Since
the outcomes have different scales, we treat them separately in our models.

@@

We then proceed to the meta-analysis. We matched the house size variables with
our measures of government spending and created a theoretical $3 \times 3$
matrix. Out of the 9 possible variable combinations, we found only 7 in
the article pool. Our sample includes no papers that analyse the relationships
between log($n$) and public expenditure per capita or between $k$ and the
logarithm of public expenditure per capita.

@@

\begin{landscape}
\begin{figure}[htb]
\begin{center}
\caption{Forest plots of the relationship between legislature size and government spending (reduced sample)}
\vspace{.3cm}
\includegraphics[width=25cm,height=17cm]{../graphs/graph1.pdf}
\label{fig:plots}
\end{center}
\end{figure}
\end{landscape}

Figure \ref{fig:plots} shows the forest plots for our restricted sample,
which includes only the `r dim(dat)[1]` main coefficients of the 
`r length(unique(dat$id))` selected papers. On the left side of the plots are 
the name of the study authors and the paper publication year. For unpublished
studies, we included the first year the paper was available online. The bars in
the middle show the reported effect sizes and the vertical lines indicate their
average, weighted by standard errors. The length of the lines represent the
precision of the estimates. The red line at the bottom of the figures displays
the aggregate coefficients plus their respective confidence intervals.

@@

@@

In a nutshell, we do not find strong evidence in favour of the "law of $1/n$".
While some models do show a positive and statistically significant result, none
of the prediction intervals are totally positive or negative. The studies also
have considerable heterogeneity, what indicates that the original coefficients
do not point consistently towards the same direction.

One reason why we have not detected any strong effects may be due to the
identification strategy authors use in their models. On the one hand, OLS and
panel data models require too many controls to make units comparable, and they
are vulnerable to omitted variable bias or post-treatment bias
[@cinelli2020making; @pearl2015conditioning]. On the other hand, estimation
methods such as instrumental variables and regression discontinuity designs
have become popular because of their high internal validity
[@angrist2008mostly]. Figure \ref{fig:plots2} shows the disaggregated effects
for two sets of models that employ causal estimation techniques. They measure
the impact of lower house size on expenditure per capita (left) and on the
natural logarithm of expenditure per capita (right).

\vspace{.06cm}

\begin{figure}[htb]
\begin{center}
\caption{Forest plots of the relationship between legislature size and government spending with regression method heterogeneity (restricted sample)}
\vspace{.3cm}
\includegraphics[width=1\linewidth, height=7.5cm]{../graphs/graph2.pdf}
\label{fig:plots2}
\end{center}
\end{figure}

@@

Papers that employed instrumental variables show some interesting variation: out
of the four coefficients listed, two are positive and two are negative. Two of
these four coefficients are significant, again one positive and one negative.
The subgroup analysis presents a SMD which is negative but insignificant. In
contrast, all papers that use regression discontinuity designs show negative and
statistically significant results. Since only three papers in our sample use
RDDs, we are cautious about predicting an overall negative relation, but they do
indicate that better identification strategies yield a zero-to-negative impact
of legislature size on expenditure, in support of the reverse "law of $1/n$".

@@

The meta-regressions show that study characteristics have a considerable
influence on reported results. Electoral system affects the relationship
between legislature size and public expenditure, but the results are not
replicable in all estimations. Publication year generates conflicting findings
in our models. Nevertheless, the meta-regressions confirm that RDDs produce
negative effects more frequently than OLS regressions.
